---
title: "Genome-Wide Association Study (GWAS)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Ensure working directory = location of this Rmd
knitr::opts_knit$set(
  root.dir = dirname(rstudioapi::getActiveDocumentContext()$path)
)
```

# GWAS Workshop

This notebook demonstrates how to run a **Genome-Wide Association Study (GWAS)** using **GAPIT in R**.

Steps covered:

-  Load required packages
-  Import phenotype data
-  Import genotype data (HapMap format)
-  Run GWAS using GAPIT
-  Visualize results


------------------------------------------------------------------------

# Install and Load Required Packages

```{r packages, message=FALSE}
packages <- c("vcfR", "dplyr")

install_if_missing <- function(pkg){
  if(!require(pkg, character.only=TRUE)){
    install.packages(pkg, dependencies=TRUE)
    library(pkg, character.only=TRUE)
  }
}

invisible(lapply(packages, install_if_missing))

library(vcfR)
library(dplyr)
```

------------------------------------------------------------------------

# Load GAPIT functions

GAPIT is a widely used GWAS pipeline.

Documentation:\
<https://zzlab.net/GAPIT/gapit_help_document.pdf>

```{r gapit}
source("http://zzlab.net/GAPIT/gapit_functions.txt")
```

------------------------------------------------------------------------

# Load Phenotype Data

Phenotype file format:

| Sample | phenotype |
|--------|-----------|
| ind1   | 10        |
| ind2   | 12        |

```{r phenotype}
pheno.dat <- read.table("../data/trait.txt", header = TRUE)
colnames(pheno.dat) <- c("Sample","phenotype")

head(pheno.dat)
```

------------------------------------------------------------------------

# Load genotype data (HapMap)

```{r genotype}
hmp <- read.delim("../data/final_data.hmp.txt", header = FALSE)
```

------------------------------------------------------------------------

# Run GWAS using GAPIT

We will test several GWAS models:

-   GLM
-   MLM
-   FarmCPU
-   BLINK We will use a kinship (relationship) matrix and a PCA (principal Component Analysis to accoun of population structure, previously calculated from the whole genome. But if you dont give this to hthe program it will calculate

```{r}
myKI <- read.csv("../data/Genotype.Kin.csv", header = FALSE)
colnames(myKI) <- c("Taxa", myKI[,1])

myCV <- read.csv("../data/Genotype.PCA.csv", header = TRUE)
```

and carry out the analysis

```{r gwas}
myGAPIT <- GAPIT(Y = pheno.dat, G = hmp, model = c("GLM","MLM","FarmCPU","Blink"),
                 PCA.total = 0, Random.model = FALSE, KI=myKI, CV=myCV)
```

------------------------------------------------------------------------

# Output Files

GAPIT automatically produces:

-   Manhattan plots
-   Q-Q plots
-   GWAS result tables

These files will be saved in the **working directory**.

------------------------------------------------------------------------

# What to infer from the analysis?

## Q-Q plots:

Good GWAS results show:

- **Points follow the diagonal initially** → Most SNPs behave as expected under the null hypothesis

- **Points deviate upward only at the end** → A few SNPs show stronger association than expected

Too much deviation early would indicate:

-   population structure
-   cryptic relatedness
-   poor model correction

### What we see in our results ?

-   Most points follow the diagonal → **good model fit**
-   Strong upward deviation at the tail → **candidate loci detected**

------------------------------------------------------------------------

## Why do different models give different results?

The analysis used four models:

| Model   | Description                |
|---------|----------------------------|
| GLM     | simple regression          |
| MLM     | controls kinship           |
| FarmCPU | iterative marker selection |
| BLINK   | improved FarmCPU           |

### What we observe

#### GLM and MLM

-   early deviation
-   detects many SNPs

Interpretation: likely more false positives

#### FarmCPU and BLINK

-   strongest significant signals
-   best balance between power and false positives

Interpretation: better performing GWAS models, with FarmCPU giving the more reliable results with these data.

------------------------------------------------------------------------

## Manhattan plot: locating candidate regions

Use the Manhattan plot to identify where are the associations in the genome

### The plot shows

-   X-axis: genomic position
-   Y-axis: −log10(p-value)

Each point = one SNP

Higher points = stronger association

The peaks in several genomic regions represent candidate loci affecting the trait, with clusters of SNPs instead of single SNPs due likely to linkage disequilibrium (nearby SNPs inherit together).

## The distribution significant markers plots: the phenotype distribution

Phenotype distribution plot shows: 

- Individuals grouped by genotype 
- Their trait values 
- Whether genotype affects the trait 

Interpretation: 
- Boxes overlap = weak association 
- Boxes separated = strong association

If the boxes are clearly different between alleles, it means this DNA marker is associated with the trait

------------------------------------------------------------------------

# Remember:

- If multiple models detect the same region there is a higher confidence in the signal.

- GWAS identifies genomic regions associated with the trait, NOT necessarily the causal gene.

# Next steps:

- identify genes near significant SNPs

- check gene annotation for candidate gene identification

- functional validation
